A data-driven model of nucleosynthesis with chemical tagging in latent space

### 2019-05-20 article/mcfa.tex
words

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated the abundances of each element as being
statistically independent, despite the nucleosynthesis literature showing that
most elements can be produced by more than one pathway.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is represented by a mixture model of multivariate gaussians
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on generated and real data, including literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that \todo{five} latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and \todo{eight} are preferred given \todo{X} abundances
in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
We find \todo{two} clusters are preferred in the \Galah\ data of 16 abundances
and 1,136 stars, which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We expect more components
would be preferred given complete abundance inventory of a larger sample of stars.
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim$$10^{7.5}$ abundance measurements.
%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
%enabling model parameter estimation in minutes given $\sim$$10^7$
%chemical abundance measurements.
\end{abstract}

@@

%Chemical tagging seeks to identify star formation events using the present
%day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2002}.
The detailed chemical abundances observable in a star's photosphere provide a
fossil record that carries with it information about where and when that star
formed \citep{Freeman;Bland-Hawthorn:2002}. While the photospheric abundances remain largely unchanged throughout
a star's lifetime \citep[however see][]{Dotter:2017,Ness:2018b}, the dynamical 
dissipation timescale of open clusters in the Milky Way disc is of order a few 
gigayears \citep{Portegies-Zwart:1998}. That makes chemical tagging an attractive 
approach to identify star formation sites long after those stars are no longer 
gravitationally bound to each other.

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Although biases arise when
only considering star clusters that are still gravitationally bound, the chemical
homogeneity of open clusters provides an empirical measure of how similar stars
would need to be before they could be tagged as belonging to the same
star formation site \citep{Mitschang:2014}. However, there are also analysis
issues in understanding how precisely those chemical abundances can be measured
\citep{Bovy:2016}, and how chemically similar stars can be which did not form 
together \citep[doppleg\"angers;][]{Ness:2018}.
If open clusters were truely chemically homogeneous then under idealistic 
assumptions our ability to chemically tag the Milky Way would depend primarily
on the precision with which we can measure those chemical abundances in stars. 
Data-driven approaches to modelling stellar spectra are
improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
needed: astronomers have not yet developed unbiased estimators of chemical
abundances that saturate the Cram\'er-Rao bound.

There are further chemical tagging obstacles to overcome once all the chemical 
abundance information in a stellar spectrum can be extracted. Most notably is a 
catalogue of precise detailed chemical abundances for a large number of stars,
where those chemical abundances trace different nucleosynthetic pathways.
This is the primary scientific goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
a stellar spectroscopic survey that uses the High Efficiency and Resolution 
Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
Milky Way, and measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). These data provide an unparalleled view on the production
of chemical elements in the Milky Way.

Given these data and the most favourable assumptions in chemical tagging 
-- that star clusters are truely chemically homogenous, and we can measure 
those abundances with infinite precision, and those abundances are 
differentiable between star clusters -- then chemical
tagging becomes a clustering problem. All clustering techniques applied to 
chemical tagging thus far have assumed that the data dimensions are independent. That is to say
that adding a dimension of [Ni/H] provides entirely new, independent information
that could not have been predicted from other elemental abundances.
Theory and observations agree that this cannot be true.
Nucleosynthetic processes produce multiple elements in varying
quantities, and the effective dimensionality of stellar abundance datasets has been shown
to be lower than the actual number of dimensions \citep{Price-Jones:2018}.
Any clustering approach that treats each new elemental abundance as an 
independent axis of information will therefore conclude with biased inferences
about the star formation history of our Galaxy.

%Alternative approaches are perhaps more challenging, as i
It is not trivial to estimate
the nucleosynthetic yields that have contributed to the chemical abundances of each star
with great confidence. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia contributions had a significant impact, but 
efforts to date have not sought to try to explain the detailed abundances of a 
single star as a contribution of yields from different systems (e.g., 20\% of the
Ba was produced by 3 M$_\odot$ asymptotic giant branch stars). This is in part 
because of the challenging and degenerate nature of the problem as described, 
and is complicated by the differences in yield predictions that account from 
prescriptions used in different codes.

New approaches to chemical tagging are clearly needed. Immediate advances would
include methods that take the dependence among chemical elements into account
within some generative model, or techniques that combine chemical abundances
with dynamical constraints to place joint prior probabilities on whether any
two stars could have formed from the same star cluster, given some model of the
Milky Way. In this work we focus on the former.

Here we present a new approach to chemical tagging that allows us to identify 
the latent (unobserved) factors that contribute to the chemical abundances of 
all stars (e.g., nucleosynthetic yields) while simultaneously performing 
clustering in the latent space. Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusions}.

@@

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
Related techniques include principal component analysis \citep{Hotelling:1933} and its
variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars \citep[e.g.,][]{Price-Jones:2018}, factor
analysis cannot describe clustering in the data (or latent) space.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
data.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, and the factor scores 
for the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from the $\numcomponents$th multivariate normal distribution.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

We assume that the latent space is lower dimensionality than the
data space (e.g., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{typical} yields
of nucleosynthetic
events (e.g., $s$-process production from AGB stars averaged over
initial mass function and star formation history), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the latent processes that are
common to all stars (the so-called factor loads, analogous to 
nucleosynthetic yields). Within this framework a rare nucleosynthetic event
can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have negligible contribution
to the observed abundances. In practice the factor loads can only be 
identified up to orthogonality and cannot be expressly interpreted as
nucleosynthetic yields because they have limited physical meaning
(we discuss this further in Section~\ref{sec:discussion}),
but this description of typical yields and relative event rates should
help build intuition for the model parameters, and provide context
within the astrophysical problem it is being applied.

@@

From a statistical standpoint, the primary advantage to using
a mixture of factor analysers is that we can simultaneously
estimate latent factors (e.g., infer nucleosynthetic 
yields) and perform clustering (e.g., chemical tagging) 
within a statistically consistent framework. That is to say
that we have a generative data-driven model that can 
quantitatively account for nucleosynthetic yields, variations in
turbulence and gas mixing, or star formation efficiency,
and the parameters of this model can be estimated consistently
given some data.

@@

Here we describe the default initialisation of the model parameters, which
we have found to be robust in a large set of experiments. Given a number of
$\NumLatentFactors$ latent factors we initialise the factor load
$\factorloads$ entries to be drawn from a uniform distribution
$\factorloads \sim \mathcal{U}\left(-1, +1\right)$ and then ensure that
the factor loads are orthogonal such that
%TODO: Should we re-write this so that we just draw from the scipy special ortho group?
\begin{equation}
	\factorloads \factorloads\transpose = \eye \quad .
\end{equation}
We then initially assign each data point as belonging to one of the
$\NumComponents$ components by drawing from a multinomial distribution
and generating the $\NumData \times \NumComponents$ responsibility matrix $\vec\tau$,
where $\sum_{k=1}^{K} \tau_{nk} = 1$ for any $n$th row. Given the initial
factor loads and assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
the covariance matrix of factor scores of each component $\scorecovs$.
Finally, we initialise the specific variance $\specificvariance$ in each
dimension as the variance in each data dimension. Other initialisation 
methods are available in the code associated with this \emph{Article}. 
Throughout this work we found that the optimised model parameters appear
to be insensitive to the choice of initialisation up until orthogonal
rotation of the latent space.

@@

We use the expectation-maximization algorithm to estimate the model parameters
\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
given the model parameters $\vec\Psi$, and we re-calculate the $\NumData \times \NumComponents$ responsibility 
matrix $\vec\tau$ whose entries are the posterior probability that the 
$\numdata$th data point is associated to the $\numcomponents$th component, given 
the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \diag{\specificvariance})} \quad .
\end{equation}

At the maximization step we update our estimates of the parameters,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this difference is inconsistency becomes serious at small $N$.}
In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations in order to ensure that the global minima is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.

The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ for each component are then given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

One criterion commonly employed for evaluating a class of models is the 
Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
\end{equation} 
\noindent{}where $Q$ is the number of model parameters, which for a mixture of
factor analyzers is:
\begin{equation}
	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
\end{equation}
\noindent{}We evaluate the BIC for all experiments performed in this work.
However, while the BIC does include a penalisation term for the number of
parameters (which scales with $\log{N}$), it does not describe for the
increased flexibility that is afforded by the addition of those parameters.
For example, adding one parameter
to a model will increase the BIC by at most $\log{N}$, but there are different
ways for a single parameter to be introduced. In a fictitious model $y=f(x)$
a parameter $b$ could be added that is a scalar multiple of $x$, or it could be
introduced as $x^b$. Despite the difference in model complexity, the same
penalisation occurs in BIC. Even if the log likelihood were only to improve
marginally in both cases, the difference in model complexity is not captured
by BIC. In other words, there are situations where we are more interested in
balancing the model complexity (or the expected Fisher information and similar
properties) with the goodness of fit, instead of penalising the number of 
parameters.

@@

The \textit{message} must encode two parts: the model, and the data given the
model. The encoding of the message is based on Shannon's information theory~\citep{Shannon:1948}. 
The information gained from an event $e$ occurring, where $p(e)$ is the
probability of that event, is $I(e) = -\log_{2}{p(e)}$. The information content
is largest for improbable outcomes, and smallest for outcomes that we are 
almost certain about. In other words, an outcome that has a probability close
to unity has zero information content because nothing new is learned from it,
whereas rarer events convey a much higher information content.

@@

\printinunitsof{in}\prntlen{\textheight}
and 
\printinunitsof{in}\prntlen{\textwidth}

@@

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. \todo{The panels on the left show the absolute entries for each factor load.} 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}

>>>>

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is represented by a mixture model of multivariate gaussians
in a lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We perform experiments on generated and real data, including literature chemical abundances
of metal-poor stars, and those from the second \Galah\ data release. 
We find that four latent factors are preferred to explain the detailed 
abundances of metal-poor stars, and five latent factors are preferred given 
16 chemical abundances in \Galah\ data. In both experiments we recover latent factors that 
are representative of mean yields expected from the rapid- and slow-neutron
capture process, as well as nucleosynthesis in Type II supernovae. 
However, in both experiments we also identify latent factors that link
the production of light (e.g., K, Al) and heavy (e.g., Ba, Eu) elements.
We find three clusters are preferred given 1,136 stars in \Galah,
two of which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
%The third is consistent
%\todo{with what.}. 
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim$$10^{7.5}$ abundance measurements.
%We release software that scales \todo{as $\mathcal{O}(XX)$}, where $X$ is \todo{what},
%enabling model parameter estimation in minutes given $\sim$$10^7$
%chemical abundance measurements.
\end{abstract}

@@

The detailed chemical abundances that are observable in a star's photosphere provide a
fossil record that carries with it information about where and when that star
formed \citep{Freeman;Bland-Hawthorn:2002}. While the photospheric abundances remain largely unchanged throughout
a star's lifetime \citep[however see][]{Dotter:2017,Ness:2018b}, the dynamical 
dissipation timescale of open clusters in the Milky Way disc is of order a few 
gigayears \citep{Portegies-Zwart:1998}. That makes chemical tagging an attractive 
approach to identify star formation sites long after those stars are no longer 
gravitationally bound to each other.

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Although biases arise when
only considering star clusters that are still gravitationally bound, the chemical
homogeneity of open clusters provides an empirical measure of how similar stars
would need to be before they could be tagged as belonging to the same
star formation site \citep{Mitschang:2014}. However, there are analysis
issues in understanding how precisely those chemical abundances can be measured
\citep{Bovy:2016}, and how chemically similar stars can be which did not form 
together \citep[doppleg\"angers;][]{Ness:2018}.
If open clusters were truely chemically homogeneous then under idealistic 
assumptions our ability to chemically tag the Milky Way would depend primarily
on the precision with which we can measure those chemical abundances in stars. 
Data-driven approaches to modelling stellar spectra are
improving upon this precision \citep{Ness:2015,Ness:2018a,Ness:2018b,
Casey:2016,Casey:2017,Ho:2017b,Ho:2017a,Leung;Bovy:2018}, but more work is
needed: astronomers have not yet developed unbiased estimators of chemical
abundances that saturate the Cram\'er-Rao bound.

%There are further chemical tagging obstacles to overcome once all the chemical 
%abundance information in a stellar spectrum can be extracted. Most notably is a 
Chemical tagging experiments require a catalogue of precise chemical abundance
measurements for a large number of stars,
where those chemical abundances trace different nucleosynthetic pathways.
This is the primary goal of the \Galah\ survey \citep{DeSilva:2015,Martell:2017,Buder:2018},
a stellar spectroscopic survey that uses the High Efficiency and Resolution 
Multi-Element Spectrograph \citep[\project{HERMES};][]{Sheinis:2015} on the Australian 
Astronomical Telescope (AAT).  \Galah\ will observe up to $10^6$ stars in the 
Milky Way, and measure up to 30 chemical abundances for each star. This includes
light odd-Z elements (e.g., Na, K), elements produced through
alpha-particle capture (e.g., Mg, Ca, Ti), and elements produced
through the slow (e.g., Ba) and rapid neutron-capture process
(e.g., Eu). No other spectroscopic survey provides an equivalent set of
chemical abundances for a comparable number of stars.

%These data provide an unparalleled view on the production
%of chemical elements in the Milky Way.

Given these data and the most favourable assumptions in chemical tagging 
-- that star clusters are truely chemically homogenous, and we can measure 
those abundances with infinite precision, and those abundances are 
differentiable between star clusters -- then chemical
tagging becomes a clustering problem. All clustering techniques applied to 
chemical tagging thus far have assumed that the data dimensions are independent. That is to say
that adding a dimension of [Ni/H] provides entirely new, independent information
that could not have been predicted from other elemental abundances.
Theory and observations agree that this cannot be true.
Nucleosynthetic processes produce multiple elements in varying
quantities, and the effective dimensionality of stellar abundance datasets has been shown
to be lower than the actual number of dimensions \citep{Ting:2012,Price-Jones:2018}.
Any clustering approach that treats each new elemental abundance as an 
independent axis of information will therefore conclude with biased inferences
about the star formation history of our Galaxy.

%Alternative approaches are perhaps more challenging, as i
It is not trivial to estimate
the nucleosynthetic yields that have contributed to the chemical abundances of each star
with great confidence. There are
qualitative statements that can be made for large numbers of stars, or particular
types of stars, but quantifying the precise contribution of different processes
to each star is an unsolved problem. For example, the so-called [$\alpha$/Fe] `knee' in
abundance ratios in the Milky Way can qualitatively be explained by 
core-collapse supernovae being the predominant nucleosynthetic process in the
early Milky Way before Type Ia supernovae had a significant contribution, but 
efforts to date have not sought to try to explain the detailed abundances of a 
single star as a contribution of yields from different systems (e.g., 20\% of the
Ba was produced by 3 M$_\odot$ asymptotic giant branch stars). This is in part 
because of the challenging and degenerate nature of the problem as described, 
and is complicated by the differences in yield predictions that account from 
prescriptions used in different theoretical models.

New approaches to chemical tagging are clearly needed. Immediate advances would
include methods that take the dependence among chemical elements into account
within some generative model, or techniques that combine chemical abundances
with dynamical constraints to place joint prior probabilities on whether any
two stars could have formed from the same star cluster, given some model of the
Milky Way.

In this work we focus on the former.
Here we present a new approach to chemical tagging that allows us to identify 
the latent (unobserved) factors that contribute to the chemical abundances of 
all stars (e.g., nucleosynthetic yields) while simultaneously performing 
clustering in the latent space. Notwithstanding caveats that we will
discuss in detail, this allows us to infer nucleosynthetic yields rather than
strictly prescribe them from models. Moreover, the scale of the clustering
problem reduces by a significant fraction because the clustering is performed in
a lower dimensional latent space instead of the higher dimensional data space.
In Section~\ref{sec:methods} we describe the model and the methods we use to
estimate the model parameters. Section~\ref{sec:experiments} describe the 
experiments performed using generated and real data sets. We discuss the results
of these experiments in Section~\ref{sec:discussion}, including the caveats with
the model as described. We conclude in Section~\ref{sec:conclusions}.

@@

Factor analysis is a common statistical approach for describing correlated 
observations with a lower number of latent variables \citep[e.g.,][]{Thompson:2004}.
Related techniques include principal component analysis \citep{Hotelling:1933} and its
variants \citep{Tipping;Bishop:1999}, singular value decomposition \citep{Golub:1970}, and other
matrix factorization methods. While factor analysis on its own is a useful
dimensionality reduction tool to identify latent factors that contribute to
the chemical abundances of stars \citep[e.g.,][]{Ting:2012,Price-Jones:2018}, factor
analysis cannot describe clustering in the data (or latent) space.
Similarly, clustering techniques applied to chemical abundances to date 
\citep[e.g.,][]{Hogg:2016} do not account for the lower dimensionality of the
data.

@@

\noindent{}where $\factorloads$ is a $\NumLatentFactors \times \NumDimensions$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata$th data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from the $\numcomponents$th multivariate normal distribution.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumData \times \NumLatentFactors$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$.

We assume that the latent space is lower dimensionality than the
data space (e.g., $\NumLatentFactors < \NumDimensions$).
Within the context of stellar abundances, the factor loads
$\factorloads$ can be thought of as the \emph{mean} yields
of nucleosynthetic
events (e.g., $s$-process production from AGB stars averaged over
initial mass function and star formation history), and the
factor scores are analogous to the relative counts of those 
nucleosynthetic events. The clustering in factor scores
achieves the same as a clustering procedure in data space,
except we simultaneously estimate the latent processes that are
common to all stars (the so-called factor loads, analogous to 
nucleosynthetic yields). Within this framework a rare nucleosynthetic event
can still be described as a `factor load' $\factorloads_\numlatentfactors$, 
but its rarity would be represented by associated factor
scores being zero for most stars and thus have negligible contribution
to the observed abundances. In practice the factor loads can only be 
identified up to orthogonality and cannot be expressly interpreted as
nucleosynthetic yields because they have limited physical meaning
(we discuss this further in Section~\ref{sec:discussion}),
but this description of typical yields and relative event rates should
help build intuition for the model parameters, and provide context
within the astrophysical problem it is being applied.

@@

From a statistical standpoint, the primary advantage to using
a mixture of factor analysers is that we can simultaneously
estimate latent factors (e.g., infer nucleosynthetic 
yields) and perform clustering (e.g., chemical tagging) 
within a statistically consistent framework. That is to say
that we have a generative data-driven model that can 
quantitatively describe nucleosynthetic yields, and the
factor scores describe the variance in turbulence and gas mixing,
or star formation efficiency, and the parameters of this model
can be simultaneously estimated in a self-consistent way.

@@

Here we describe the default initialisation of the model parameters, which
we have found to be robust in a large set of experiments. Given a number of
$\NumLatentFactors$ latent factors we initialise the factor load
$\factorloads$ entries to be drawn from a uniform distribution
$\factorloads \sim \mathcal{U}\left(-1, +1\right)$ and then ensure that
the factor loads are orthogonal such that
%TODO: Should we re-write this so that we just draw from the scipy special ortho group?
\begin{equation}
	\factorloads \factorloads\transpose = \eye \quad .
\end{equation}
We then initially assign each data point as belonging to one of the
$\NumComponents$ components using the \texttt{k-means++} algorithm \citep{Arthur:2007}.
During the Experiments described in Section~\ref{sec:experiments} we repeat
this initialisation procedure 25 times for every permutation of $\NumLatentFactors$ and $\NumComponents$. We run expectation-maximization (Section~\ref{sec:expectation-maximization})
from each initialisation until convergence and adopt the model with the
highest log likelihood as the preferred model for that permutation of $\NumLatentFactors$ and $\NumComponents$.

Given the initial
factor loads and assignments, we then estimate the relative weights
$\vec\pi$, the mean factor scores of each component $\scoremeans$, and
the covariance matrix of factor scores of each component $\scorecovs$.
Finally, we initialise the specific variance $\specificvariance$ in each
dimension as the variance in each data dimension. Other initialisation 
methods are available in the code associated with this \emph{Article}. 
%Throughout this work we found that the optimised model parameters appear
%to be insensitive to the choice of initialisation up until orthogonal
%rotation of the latent space.

@@

We use the expectation-maximization algorithm to estimate the model parameters
\citep{Dempster:1977}. With each expectation step we evaluate the log likelihood 
given the model parameters $\vec\Psi$ and we calculate the $\NumData \times \NumComponents$ responsibility 
matrix $\vec\tau$ whose entries are the posterior probability that the 
$\numdata$th data point is associated to the $\numcomponents$th component, given 
the data $\vecdata$ and the current estimate of the parameter vector $\vec\Psi$:
\begin{equation}
	\tau_{\numdata\numcomponents} = \frac{\weight_\numcomponents\phi(\vecdata_\numdata;\factorloads\scoremeans_\numcomponents, \factorloads\scorecovs_\numcomponents\factorloads\transpose + \diag{\specificvariance})}{\sum_{g=1}^{G}\weight_g\phi(\vecdata_\numdata;\factorloads\scoremeans_g, \factorloads\scorecovs_g\factorloads\transpose + \diag{\specificvariance})} \quad .
\end{equation}

At the maximization step we update our estimates of the parameters $\vec\Psi$,
conditioned on the data $\vecdata$ and the responsibility matrix $\vec\tau$.
The updated parameters estimates are found by setting the second derivative
of the log likelihood (Eq.~\ref{eq:log-likelihood}) to zero and solving for
the parameter values\footnote{Strictly this introduces a statistical inconsistency in that we should update our parameter estimates by setting the second derivative of our chosen objective function (Eq.~\ref{eq:message-length}) to zero instead of the log likelihood, but this inconsistency only becomes serious with small $N$.}
In doing so this guarantees that every updated
estimate of the model parameters is guaranteed to increase the log likelihood.
Although there are no guarantees against converging on local minima, in 
practice it is sufficient to run expectation-maximization from multiple
initialisations (as we do) in order to ensure that the global minima is reached.
At the maximization step we first update our estimate of the relative weights 
$\vec\weight\nextstep$ given the responsibility matrix $\vec\tau$
\begin{equation}
	\weight_\numcomponents\nextstep = \frac{1}{\NumData} \sum_{\numdata=1}^{\NumData}\tau_{\numdata\numcomponents}
	% Include proof.
\end{equation}
\noindent{}where the $\vec{\Psi}\thisstep$ superscript refers to the current parameter estimates and $\vec{\Psi}\nextstep$ refers to the updated estimate for the next iteration.
The updated estimates of the mean factor scores 
$\scoremeans\nextstep$ for each component are then given by
\begin{eqnarray}
	\scoremeans_\numcomponents\nextstep = \scoremeans_\numcomponents\thisstep + \frac{\vec{G}\transpose(\vecdata\transpose - \factorloads\thisstep\scoremeans_\numcomponents\thisstep)\vec\tau_\numcomponents}{\NumData\weight_\numcomponents\nextstep}
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
	\vec{W} &=& (\scorecovs_\numcomponents\thisstep)^{-1}\eye \\
	\vec{V} &=& \left(\specificvariance\thisstep\right)^{-1} \\
	\vec{C} &=& (\vec{W} + (\factorloads\thisstep)\transpose\vec{V}\factorloads\thisstep)^{-1}\eye \\
	\vec{G} &=& \left[\vec{V} - \vec{V}\factorloads\thisstep\vec{C}\left(\vec{V}\factorloads\thisstep\right)\transpose\right]\factorloads\thisstep\scorecovs_k\thisstep \quad .
\end{eqnarray}

@@

One criterion commonly employed for evaluating a class of models is the 
Bayesian Information Criterion \citep[BIC;][]{Schwarz:1978}, 
\begin{equation}
	\textrm{BIC} = Q\log{N} - 2\log\mathcal{L}\left(\data|\vec\Psi\right) \quad , \label{eq:bic}
\end{equation} 
\noindent{}where $Q$ is the number of model parameters:
\begin{equation}
	Q = \frac{\NumLatentFactors}{2}\left[2\left(\NumDimensions - \NumLatentFactors\right) + \NumComponents\left(3 + \NumLatentFactors\right)\right] + \NumComponents + \NumDimensions - 1 \quad .
\end{equation}
\noindent{}We evaluate the BIC for all experiments performed in this work.
However, while the BIC does include a penalisation term for the number of
parameters (which scales with $\log{N}$), it does not describe for the
increased flexibility that is afforded by the addition of those parameters.
For example, adding one parameter
to a model will increase the BIC by at most $\log{N}$, but there are different
ways for a single parameter to be introduced. In a fictitious model $y=f(x)$
a parameter $b$ could be added that is a scalar multiple of $x$, or it could be
introduced as $x^b$. Despite the difference in model complexity, the same
penalisation occurs in BIC. Even if the log likelihood were only to improve
marginally in both cases, the difference in model complexity is not captured
by BIC. In other words, there are situations where we are more interested in
balancing the model complexity (or the expected Fisher information and similar
properties) with the goodness of fit, instead of penalising the number of 
parameters.

@@

The \textit{message} must encode two parts: the model, and the data given the
model. The encoding of the message is based on Shannon's information theory~\citep{Shannon:1948}. 
The information gained from an event $e$ occurring, where $p(e)$ is the
probability of that event, is $I(e) = -\log_{2}{p(e)}$. The information content
is largest for improbable outcomes, and smallest for outcomes that we are 
almost certain about. In other words, an outcome that has a probability close
to unity has nearly zero information content because almost nothing new is learned from it,
whereas rarer events convey a much higher information content.

@@

@@

\vspace{2em}

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{experiments/exp2-\ExperimentHash-latent-factors-and-contributions.pdf}
	\caption{Latent factors found in Experiment~2 using the \citet{Barklem:2005}
			 data. The panels on the left show the absolute entries for each factor load. 
			 On the right we show the absolute fractional contributions to
			 each element, illustrating the loads that contribute most to each element.}
    \label{fig:exp2-factor-loads}
\end{figure*}
